Estimating duration in the presence of misclassification

### 2024-01-26 transmission.tex
Overleaf editing

In this chapter, I use the preferred estimate of the duration of PCR positivity (see \cref{E-imperf-test:sec:conclusion}) to estimate the transmission of SARS-CoV-2 in England.
This estimation is based on the CIS data (see \cref{transmission:sec:data} for further details).
Estimating transmission is the ultimate aim of this thesis.

I take two approaches; the theoretical advantages of each were discussed in \cref{E-inc-prev:sec:infection-process}.
First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
Then, in \cref{SEIR}, I use a mechanistic compartmental model, an implementation of the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.

@@

As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
Therefore, I extend the methodology previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
Here, a Bayesian multilevel logistic regression model is first fit to predict PCR positivity from demographic covariates.
Then, the posterior distribution is used to predict the prevalence in each stratum.
Finally, the prevalence is poststratified to the CIS target population and subpopulations of interest.
I follow this same process, except I transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before poststratifying.
% I fit a modification model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
% Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.

@@

I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
First, a Gaussian Markov Random Field (GMRF) logistic regression model is used to estimate the prevalence in each stratum.
For a day $t$, age group $a$, ethnicity $e$, and sex $s$ assume that:
\begin{align}
    \logit(p_{taes}) &= \beta + \beta_s + \beta_{w(e)} + u_a + v_e + w_t + l_{a,t} \\
    y_{taes} &\dist \BB(n_{taes}, p_{taes}, \rho)
\end{align}
The terms are as follows.
\begin{itemize}
    \item $\beta$ is a global intercept term with the prior $\beta \dist \N(0, 1/0.001^2)$.
    \item $\beta_s$ is a fixed effect for sex, with $\beta_s = 0$ for males and estimated for females. $\beta_{w(e)}$ is a fixed effect for white or non-white ethnicity. $\beta_{w(e)} = 0$ if $e$ corresponds to white ethnicity and $\beta_{w(e)} = \beta_{n}$ otherwise (\ie non-white ethnic groups).% $\beta_n$ and $\beta_f$ have the default INLA prior for fixed effects\todo{check what this is}.
    \item $u_a$ and $v_e$ are independent Gaussian random effects for age and ethnic group respectively. Their standard deviation have a penalised complexity prior\todo{cite PC prior} with a 10\% probability of being greater than 1.
    \item $w_t$ is a second-order random walk, that is $w_t = 2w_{t-1} - w_{t-2} + \epsilon_t$ where $\epsilon_t \dist \N(0, \sigma_w^2)$.
    \item $l_{a,t}$ is an age-group specific second-order random walk. That is $l_{a,t} = 2l_{a,t-1} - l_{a,t-2} + \epsilon_{a,t}$ where $\epsilon_{at} \dist \N(0, \sigma_l^2)$.
    \item The likelihood is a beta-binomial model (defined in \cref{E-distributions}), an overdispersed binomial. Overdispersion allows for household clustering\todo{cite beta-binomial as a model for clustering} and other overdispersion in the data, which is common in epidemiological data\todo{cite that epi data is often overdispersed}. 
\end{itemize}
All parameters have priors using the default specified in INLA, except where stated above.
For these parameters, the large amount of data means that the prior has little influence on the posterior distribution.

The model was fit separately and independently for each region, implicitly adding interactions between region and all other variables.
This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.

Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies.
Therefore, \cref{E-inc-prev:eq:EPt} implies that $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix of the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
The matrix $\matr{S}$ is defined to maintain the relationship between incidence and prevalence in \cref{E-inc-prev:eq:EPt}.
This gives a lower-triangular of the following form:
\begin{align}
    (\matr{S})_{i,j} = \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases}
\end{align}
Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
Forward substitution is numerically stable unless the diagonal elements are close to 0.
Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
\todo{I explained this more clearly in my first-year report, look that up!}

\todo{Insert constant incidence assumption methodology}

Model misspecification means that the procedure described could produce negative incidence.
For any stratum and day $t$, this occurs if the difference between $P_t$ and $P_{t+1}$ is too large based on the number of recoveries that occur on day $t$ based on the incidence up to day $t$ and the duration distribution.

There are two places where this misspecification could occur.
The first is the survival matrix, $\matr{S}$.
This is assumed known but \cref{E-imperf-test} showed that it is sensitive to assumptions made.
The second is the model used to estimate prevalence.

A prior constraining incidence to be non-negative and then fit directly to the data could be used but, as previously discussed, this would no longer be a GLM and could not be fit within the R-INLA framework.
Hence, an alternative inference methodology would be required.
MCMC is likely too computationally expensive.
Other approaches, such as a Laplace approximation or variational inference, could be considered.
For this analysis, I replaced any value of incidence estimated to be negative with 0.

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
For each posterior draw and day, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
The population numbers for each stratum are provided by the Office for National Statistics.

\subsection{Results and discussion} \label{backcalc:sec:results}

A slow increase in incidence occurs over the autumn 2020 period, peaking in late November in most regions (see \cref{transmission:fig:backcalc-regions}).
An exception is the North East and East of England regions (see \cref{transmission:fig:backcalc-regions}(C) and (D) respectively); these regions show no consistent increase or decrease in incidence during October.
All regions then decrease during November, coinciding with the national lockdown.
A rise then occurs, although starting at different times in different regions.
This could be due to the emergence of the Alpha variant, which was more transmissible than previous variants and emerged at different times across the country\todo{cite Alpha variant}.
This leads to a second peak in late December or early January in all regions; this second peak is larger than the first in all regions except the Northern three (North East, North West, and Yorkshire and the Humber).
\begin{figure}
    \includegraphics{transmission/backcalc-regions}
    \caption[Incidence estimated using backcalculation by region]{%
        Incidence proportion for all of England (A) and its component regions.
        Regions are split into broad groupings of three regions each: Northern (B), Midlands (C), and Southern (D).
        Yorkshire and the Humber is labelled as Yorkshire for brevity.
    }
    \label{transmission:fig:backcalc-regions}
\end{figure}

The incidence proportion was highest among the 11--15 and 16--25 age groups (see \cref{transmission:fig:backcalc-ages}).
Except the 16--25 age group, all age groups show similar trends (\ie parallel lines on a log scale).
This is the pattern predicted by an age-stratified transmission model (see \cref{E-SEIR:sec:structured-populations}), if behavioural changes are independent of age group, and there is negligible impact of immunity.
The 16--25 age group decreases faster during November.
There are many plausible explanations for this effect.
The higher incidence in this age group prior to November means that this could be an immunity effect.
Alternatively, it could be a consequence of schools remaining open in the November lockdown, which has a small effect on 16--25 as they interact less with school-age children than other age groups.
Many other explanations are possible, and further work is required to determine causality.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{transmission/backcalc-ages}}
    \caption[Incidence estimated using backcalculation by age group]{%
        Incidence proportion for each age group, on both natural and log scales.
    }
    \label{transmission:fig:backcalc-ages}
\end{figure}

An edge effect where a rapid rise is seen between day 1 and 2 is visible (see \cref{transmission:fig:start-effect}).
This is due to the incorrect assumption that incidence is constant in the past.
Changing this assumption (\eg to exponential growth) modifies the estimated value on the first day, but the rest of the series is insensitive to this change (not shown).
Therefore, the first day's estimate is unreliable, and has been excluded from all other plots.
However, the rest of the series is robust to this assumption and hence reliable.
\begin{figure}
    \centering \includegraphics{transmission/backcalc-start-effect}
    \caption[Edge effects in backcalculation method]{%
        First seven days of incidence proportion estimate for the whole of England, highlighting the edge effect.
    }
    \label{transmission:fig:backcalc-start-effect}
\end{figure}

@@

Studies for pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

@@

The modelled proportion of individuals that are PCR positive in the are linked to the data using a beta-binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
In this parameterisation (detailed in \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.

@@

    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
\end{tikzpicture}
}
  \caption[SEIR model]{SEIR model used for the analysis in this section. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
  \label{SEIR:fig:full-model}
\end{figure}

@@

Priors for the parameters are given in \cref{SEIR:table:priors}.
The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors are used for them.
$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
The prior on $\rho$ favours a null model of a binomial likelihood.
The remaining parameters had weakly informative priors.

@@

In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.

For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value (see \cref{SEIR:fig:true-vs-posterior}).
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
\end{figure}

@@

The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
Incidence is smooth within each week but shows discontinuities between weeks.
This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
When the peak in incidence is estimated to occur varied by region, with the most common time being early January 2021 (see \cref{SEIR:fig:peak-incidence}).
\begin{figure}
    \includegraphics{SEIR/CIS/incidence}
    \caption[Posterior estimates of incidence]{%
        Posterior estimates (95\% CrI) of incidence by region and age.
        See main text for discussion.
    }
    \label{SEIR:fig:incidence}
\end{figure}
\begin{figure}
    \includegraphics{SEIR/CIS/p_peak}
    \caption[Posterior estimates of peak incidence timing]{%
        Posterior probability of the peak incidence for each region and age combination occurring on each day.
    }
    \label{SEIR:fig:peak-incidence}
\end{figure}

@@

The \emph{attack rate} is the proportion of the population that have been infected at least once\todo{cite attack rate}.
In this model it is $1 - \vec{s}$, the proportion susceptible.
The attack rate is highest in the $[11, 16)$ age group in the North West at 68\% (95\% CrI: 66\%--70\%) (see \cref{SEIR:fig:attack-rates}).
\begin{figure}
    \includegraphics{SEIR/CIS/attack_rates}
    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
    \todo[inline]{Maybe compare this to blood donors, as used in the prior.}
    \label{SEIR:fig:attack-rates}
\end{figure}

Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
There are a few features of the data which are not captured by the model and the posterior estimate appears to be a compromise with the prevalence too low in one age group and too high in another.
For example, in the North East region, the 11--16 age group has its prevalence overestimated in October but the 16--25 age group has its prevalence underestimated.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev-young}
\end{figure}
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
    }
    \label{SEIR:fig:prev-old}
\end{figure}

All regions' MCMC chains converged, assessed using the Rhat statistic and ESS (as defined in \cref{E-inc-prev:sec:MCMC}).
The lowest Rhat was 1.01 for $\sigma_\epsilon$ in London.
All parameters had ESS > 1000, except $\sigma_\epsilon$ and $\psi$ in London, which had ESS of 450 and 950 respectively.

@@

The results in this section show the feasibility of fitting a mechanistic model using only data on prevalence.
If the distributions for length-of-stay in each compartment are known, then all other model parameters can  be estimated.
This includes the age distribution of the infections and the susceptibility of children.

@@

The parameters $\beta_w$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
Many factors contribute to the pattern in the $\beta_w$ values, which have a minima near the start of December.
While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.

The decreases in $\beta_w$ over the autumn could be due to increased caution (\eg masking or stricter adherence to social distancing) as cases and deaths increased~\autocite{jarvisEffect} and/or interventions not captured by $\matr{\kappa}$ such as the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
Local measures are not well captured by $\matr{\kappa}$ because the matrix averages behaviour across the whole of England.
Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately, although their targeted nature may require more granular methods.

The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite[e.g.][]{daviesEstimated}.
A multiple variant model might be able to capture this rise more accurately.
Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
Then, each variant would be explicitly included in the model to fit to this data.

@@

The current model could be more flexible with regards to the age distribution of infections.
This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
This implies that the data would allow identification of a more flexible age-based model.

Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $q_c$.
Increasing the number of parameters in $\matr{\beta}$ would be the simplest approach to increase the model's flexibility.

The only time-changing part of $\matr{\beta}$ that varies between ages is $\matr{\kappa}$.
It is plausible that this would not capture all behavioural changes.
For example, the introduction and relaxation of control measures within schools.
$\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants~\autocite{zhuRole}, which could again motivate these changing over time.

The weekly patterns in incidence are a limitation of the current parameterisation.
The random walk and contact matrices both vary only on the weekly timescale.
A daily random walk could provide more flexibility here.
However, this would greatly increase the number of parameters.
An alternative would be a smoothly-varying function or an approximation of a random walk, such as that proposed by \textcite{ghoshApproximate}.

>>>>

In this chapter, I use the estimate of the duration of PCR positivity from \cref{E-imperf-test} to estimate the transmission of SARS-CoV-2 in England.
This estimation is based on the CIS data (see \cref{transmission:sec:data} for further details).
Estimating transmission is the ultimate aim of this thesis.

I take two approaches.
The theoretical advantages of each approach were discussed in \cref{E-inc-prev:sec:infection-process}.
First, in \cref{backcalc}, I use a back-calculation approach, based on \cref{E-inc-prev:eq:EPt}.
Then, in \cref{SEIR}, I use a mechanistic compartmental model, extending the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.

@@

\todo[inline]{make sure to proofread this section due to merge conflicts}
As discussed in\todo{ref section}, a major issue with the CIS is non-response bias.
However, a deterministic relation between true prevalence and incidence is a good approximation (see \cref{E-inc-prev:sec:prevalence-process}).
Therefore, I extend the methodology previously previously used to estimate prevalence from the CIS data, corrected for non-response bias, to estimate incidence\todo{cite Koen's stuff}.
The previous method was multilevel regression and poststratification (MRP)\todo{cite MRP}.
Here, a Bayesian multilevel logistic regression model is first fit to predict PCR positivity from demographic covariates.
Then, the posterior distribution is used to predict the prevalence in each stratum.
Finally, the prevalence is poststratified to the CIS target population and subpopulations of interest.
I follow this same process, except I transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence before post-stratifying.
% I fit a modification model to estimate prevalence, then transform the posterior distribution of prevalence in each stratum to a posterior distribution of incidence.
% Finally, I poststratify the posterior distribution of incidence to the CIS target population and subpopulations of interest.

@@

I modified a pre-existing version of a MRP model for CIS~\autocite{pouwelsMRPvaccination}, as recommended by that model's author for this situation~\citePersonalComms{Koen Pouwels}.
\todo[inline]{Describe the model precisely, see eg RTM paper 2}
% The model included fixed effects for age group\todo{check the model I used}.
% The likelihood is a beta-binomial distribution, a generalisation of the binomial distribution allowing for overdispersion (see \cref{E-distributions}).
% The overdispersion allows for household clustering\todo{ref beta-binomial as a model for clustering} and any other overdispersion in the data\todo{ref that epi data is often overdispersed}. 
Finally, the model was fit independently for each region, implicitly adding interactions between region and all other variables.
This step assumes that the estimates for each region can be considered independent, which is reasonable because the regions are large with many tests conducted within each.

Having fit the prevalence model, I draw 10,000\todo{check number} prevalence trajectories in each stratum from the approximate posterior distribution of the model.
For each stratum, I use the relationship in \cref{E-inc-prev:eq:EPt} to calculate the incidence proportion at each time point.
I consider the relationship between incidence and prevalence within a stratum to be deterministic, as \cref{E-inc-prev:sec:prevalence-process} justifies.
Therefore, \cref{E-inc-prev:eq:EPt} implies that $\vec{P} = \matr{S} \vec{Z}$, where: $\vec{P}$ is a vector of the number of prevalent individuals at each time point in the stratum; $\matr{S}$ is a matrix of the survival function, as defined below; and $\vec{Z}$ is a vector of the incidence at each time point in the stratum.
The matrix $\matr{S}$needs to bebased on preserves the relationship between incidence and prevalence.
This gives a lower-triangular of the following form:
\begin{align}
    (\matr{S})_{i,j} = \begin{cases}
        S(i - j + 1) & i - \dmax - 1 \leq j \leq i \\
        0 & \text{otherwise.}
    \end{cases}
\end{align}
Since the matrix is lower-triangular, the solution for $\vec{P}$ can be found efficiently using forward substitution\todo{cite forward substitution}.
Forward substitution is numerically stable unless the diagonal elements are close to 0.
Here, the diagonal is $S(1) = 1$ and hence the matrix is well-conditioned.
\todo{I explained this more clearly in my first-year report, look that up!}

The final step is poststratifying the incidence proportion to the CIS target population and subpopulations of interest.
For each posterior draw and day\todo{introduce discrete time in days earlier}, the incidence proportion in the (sub)population is the weighted average of the incidence proportion in each stratum, where the weights are the proportion of the population in each stratum.
The population numbers for each stratum are provided by the Office for National Statistics.

\subsection{Results} \label{backcalc:sec:results}

\subsection{Discussion} \label{backcalc:sec:discussion}

@@

Studies of pre-Alpha SARS-CoV-2 suggest that all individuals have similar infectiousness but that pre-adolescent children are less susceptible than adults~\autocites{chenRole}{vinerTransmission}.
Therefore, I set $q^\text{inf}_{a'} = 1$ for all $a'$, $q^\text{susc}_{a} = q_c$ for the youngest stratum, and $q^\text{susc}_{a} = 1$ for the remaining strata.
$q_c$ is then estimated, although using an informative prior (see \cref{SEIR:sec:priors}).

@@

The modelled proportion of individuals that are PCR positive in the are linked to the data using a Beta-Binomial likelihood (see\todo{ref section on clustering} for the issues with using a binomial likelihood).
If on each day $t$, each strata $i$ is observed to have $y_{it}$ positives out of $n_{it}$ tests, then the likelihood is:
\begin{align}
    \prod_{a,t} \BB (y_{at} \mid n_{at}, \sum_{k=1}^l \bar{p}_{kat}, \rho)
    \label{SEIR:eq:likelihood}
\end{align}
where $\bar{p}_{kat} = \int_{t}^{t+1} p_{ka}(s) ds$ is the mean occupancy in compartment $p_k$ and strata $a$ over day $t$ and $\rho$ is a nuisance parameter controlling the overdispersion of the beta-binomial distribution.
In this parameterisation (detailed in \cref{E-distributions}) at $\rho=0$ the beta-binomial coincides with the binomial distribution and as $\rho \to \infty$, the distribution tends towards the uniform distribution on $[0, 1]$~\autocite{hughesUsing}.

@@

    \node[draw, thick, inner sep=0.3cm, fit=(P1) (P3), fill=blue!10,opacity=0.2] {};
\end{tikzpicture}
}
  \caption[SEIR model]{SEIR model used for the analysis in this chapter. The shaded box represents PCR-positive individuals. Shown for a single stratum.}
  \label{SEIR:fig:full-model}
\end{figure}

@@

Priors for the parameters are given in \cref{SEIR:table:priors}.
The parameters governing the duration of positivity ($\kappa$s) were explained previously (see \cref{SEIR:table:ec-params}).
The initial conditions and transition rate parameters are unindentifiable in a SEIR model~\autocite{dankwaStructural}, therefore strongly informative priors are used for them.
$\sigma_\epsilon$ and $\rho$ are given priors that penalise model complexity.
The prior on $\sigma_\epsilon$ is a penalised complexity prior, favouring a null model of constant values for $\beta(t)$~\autocite{simpsonPenalising}.
The prior on $\rho$ favours a null model of a binomial likelihood.
The remaining priors had weakly informative priors.

@@

\todo[inline]{What happened to the following paragraph? It doesn't make sense currently!}
In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown\todo{consider showing prior/posterior plots}) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \makebox[\textwidth][c]{\includegraphics{SEIR/sim/true_vs_posterior}}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
\end{figure}

@@

The 11--16 age group (roughly corresponding to compulsory secondary school education) is estimated to have the highest incidence (see \cref{SEIR:fig:incidence}).
Incidence is smooth within each week but shows discontinuities between weeks.
This is because the parameters within $\matr{\beta}(t)$ are constant within each week.
A constant $\matr{\beta}(t)$ would lead to exponential growth following a period of transitory dynamics, although the weekly switching may mean that these transitory dynamics never end, especially when incidence is high~\autocite{rhodesConvergence}.
When the peak in incidence is estimated to occur varied by region, but due to the switching, is almost always on a Sunday (see \cref{SEIR:fig:peak-incidence}).
For Yorkshire, the peak is estimated to occur in late October.
For all other regions, it is estimated to occur in late December or Early January, the period during which severe social distancing measures were applied.
\begin{figure}
    \includegraphics{SEIR/CIS/incidence}
    \caption[Posterior estimates of incidence]{%
        Posterior estimates (95\% CrI) of incidence by region and age.
        See main text for discussion.
    }
    \label{SEIR:fig:incidence}
\end{figure}
\begin{figure}
    \includegraphics{SEIR/CIS/p_peak}
    \caption[Posterior estimates of peak incidence timing]{%
        Posterior probability of the peak incidence for each region and age combination occurring on each day.
    }
    \label{SEIR:fig:peak-incidence}
\end{figure}

@@

The \emph{attack rate} is the proportion of the population that have been infected at least once\todo{cite attack rate}.
In this model it is $1 - \vec{s}$.
The attack rate is highest in the $[11, 16)$ age group in the North West at 50\% (95\% CrI: 48\%--52\%) (see \cref{SEIR:fig:attack-rates}).
\begin{figure}
    \includegraphics{SEIR/CIS/attack_rates}
    \caption{Posterior median and 95\% CrI of the attack rate in each region and age group.}
    \todo[inline]{Maybe compare this to blood donors, as used in the prior.}
    \label{SEIR:fig:attack-rates}
\end{figure}

Overall the model fits the data well (see \cref{SEIR:fig:prev-young,SEIR:fig:prev-old}).
There are a few features of the data which are not captured by the model.
In particular, prevalence during October is higher in the 16--25s than the model predicts and lower in the 11--16s.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_young}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (younger ages)]{%
        Goodness-of-fit to prevalence data by region and age for the younger three age groups (see \cref{SEIR:fig:prev-old} for the others).
        The central ribbon and line shows the posterior median and 95\% CrI estimate for the proportion of the relevant strata PCR positive on each day.
        The lighter outer ribbon is the 95\% posterior predictive interval for the proportion of test results that are positive each week.
        The dots show the observed prevalence in CIS, aggregated by week.
        For a well-calibrated model, 95\% of the dots would be within the outer ribbon.
    }
    \label{SEIR:fig:prev-young}
\end{figure}
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/CIS/prev_old}}
    \captionsetup{width=0.8\paperwidth}
    \captionof{figure}[SEIR prevalence goodness-of-fit (older ages)]{%
        Goodness-of-fit prevalence for the older three age groups (see \cref{SEIR:fig:prev-young} for the other age groups and details).
    }
    \label{SEIR:fig:prev-old}
\end{figure}

One MCMC chain for the North West region appeared to get stuck, with very few moves after the first 100,000 iterations (not shown).
I reran this chain with a new seed, with the rerun results presented.
Following this, all regions' MCMC chains converged, assessed using the Rhat statistic and ESS.
Rhat for all parameters is < 1.01, except $\sigma_\epsilon$ in London which had a Rhat of 1.018.
All parameters had ESS > 1000, except $\sigma_\epsilon$ in London and the East, which had ESS of 197 and 757 respectively.
As all the parameters of primary interest had high Rhat and ESS, and no parameters had large issues, the inference procedure is considered to have converged.

@@

The results in this chapter show the feasibility of fitting a mechanistic model using only data on prevalence.
If the distributions for length-of-stay in each compartment are known, then all other model parameters can  be estimated.
This includes the age distribution of the infections and the susceptibility of children.

@@

The parameters $\beta_w$ absorb all the time-varying factors not captured by Google mobility and school attendance data.
Many factors contribute to the pattern in the $\beta_w$ values, which have a minima near the start of December.
While this analysis cannot establish causation, several hypotheses can be identified that are worth further investigation.
The decreases in $\beta_w$ over the autumn could be due to increased caution(\eg masking or stricter adherence to social distancing) and/or interventions short of a lockdown as cases and deaths increased~\autocite{jarvisEffect}, or the impact of local measures in the highest transmission areas~\autocite{scottCovid19}.
Local measures are not well captured by $\matr{\kappa}$ because the matrix was for the whole of England.
Allowing each region to have its own $\matr{\kappa}$ may capture the effect of local measures more accurately.
The increases in $\beta_w$ in December coincide with the rise of the Alpha variant~\autocite{lythgoeLineage}, known to be more transmissible~\autocite{daviesEstimated}.
A multiple variant model might be able to capture this rise more accurately.
Such a model would stratify positive tests in the CIS by pre-Alpha or Alpha variant.
Then, each variant would be explicitly included in the model to fit to this data.

@@

The current model could be more flexible with regards to the age distribution of infections.
This would better exploit one of the strengths of the CIS: the availability of data  on all age groups.
In some regions, there appears to be a trade-off between the 11--16s and 16--25s, with the model misfitting in opposite directions (see \cref{SEIR:fig:prev-young}).
This implies that the data would allow identification of a more flexible age-based model.
Currently, the age distribution of infections is determined by $\matr{\beta}$, and to a lesser extent $\vec{\pi}$, the former of which only contains one parameter that affects the age distribution, $\beta_c$.
More parameters could capture the age distribution more accurately.
There are various ways to increase the model's flexibility, including introducing additional susceptibility or infectiousness parameters, or by allowing these parameters to vary over time.
It is plausible that these distributions changed over time in a way not captured by $\matr{\kappa}$, for example due to the introduction and relaxation of control measures within schools.
$\matr{\kappa}$ only captures changes in in-school transmission through changes in school attendance, not incorporating the probability of transmission, which could be modelled as changes in $q_{aa}$ for $a$ of school age.
Similarly, there is evidence that $q_c$ is different for Alpha and pre-Alpha variants~\autocite{zhuRole}, which could again motivate these changing over time.

The weekly patterns in incidence are another limitation of the current parameterisation.
The random walk and contact matrices both vary only on the weekly timescale.
A daily random walk could provide more flexibility here.
However, this would greatly increase the number of parameters.
An alternative would be a smoothly-varying function or an approximation of a random walk such as that proposed by \textcite{ghoshApproximate}.
